---
title: "README"
author: "Alexander Rogers"
date: "2024-03-08"
output: html_document
---

```{r}
library("tidyverse")
library("stringr")
```


   
"Hoy"
"las"
"niñas"         
"lanza"         
"pelotas"       
"pista"         
"Por"
"tarde"         
"bebe"          
"un"
"zumo"          
"comedor"       
"odia"          
"niña"          
"por"
"travesuras"    
"Dicen"         
"ayer"          
"hombre"        
"come"          
"pollo"         
"bar"
"Explican"      
"anteayer"      
"cocinero"      
"grita"         
"receta"        
"a"
"camarero"      
"chico"         
"busca"         
"chica"         
"parque"        
"mañana"        
"mejoraron"     
"plato"         
"para"          
"almuerzo"      
"los"
"hombres"       
"firmaron"      
"contrato"      
"reunión"       
"hermana"       
"gasta"         
"dinero"        
"zapatos"       
"nuevos"        
"periodista"    
"imprime"       
"noticia"       
"noche"         
"apagó"         
"luces"         
"de" 
"Normalmente"   
"abogada"       
"llama"         
"desde"         
"En" 
"este"          
"instante"      
"abuelos"       
"cambian"       
"rueda"         
"su" 
"coche"         
"hija"          
"aprende"       
"idiomas"       
"con"
"amigas"        
"llena"         
"gasolina"      
"diesel"        
"gritaron"      
"respuesta"     
"empleada"      
"primo"         
         
"pelo"          
"piso"          
"Avisan"        
"año"
"pasado"        
"todo"          
"papel"         
"caro"          
"semana"        
"pasada"        
"factura"       
"maletas"       
"unas"          
"drogas"        
"investiga"     
"caso"          
"carcel"        
"Piensan"       
"investigó"     
"secuestro"     
"policía"       
"jefe"          
"mejoró"        
"condiciones"   
"mes"
"cocinó"        
"una"
"tortilla"      
"fiesta"        
"compañía"      
"mandó"         
"cartas"        
"correo"        
"normal"        
"abuela"        
"comprende"     
"al" 
"abuelo"        
"durante"       
"peleas"        
"recibió"       
"regalo"        
"facturó"       
"guitarra"      
"aeropuerto"    
"adivinó"       
"nombre"        
"señora"        
"esperan"       
"vuelo"         
"guardan"       
"ropa"          
"armario"       
"recoge"        
"libros"        
"biblioteca"    
"gerente"       
"escala"        
"montaña"       
"hermano"       
"abraza"        
"niño"          
"tienda"        
"graban"        
"canción"       
"mujer"         
"besa"          
"cama"          
"Anuncian"      
"cancela"       
"mal"
"tiempo"        
"bombero"       
"fuego"         
"iglesia"       
"De" 
"repente"       
"nieto"         
"firma"         
"carta"         
"cirujano"      
"necesita"      
"tijeras"       
"operación"     
"barrió"        
"suelo"         
"nieta"         
"novio"         
"graba"         
"grabadora"     
"compartieron"  
"helado"        
"alumna"        
"mira"          
"profesora"     
"aula"          
"momento"       
"levantan"      
"caja"          
"Creen"         
"espera"        
"metro"         
"estación"      
"Comentan"      
"hijo"          
"cepilla"       
"perro"         
"sigue"         
"cocina"        
"niños"         
"llenó"         
"vaso"          
"nietas"        
"escucharon"    
"cuento"        
"tío"
"limpió"        
"calle"         
"necesitaron"   
"ayuda"         
"guarda"        
"aleman"        
"academia"      
"chicos"        
"mandaron"      
"novia"         
"maestro"       
"compartió"     
"respuestas"    
"anoche"        
"alumnos"       
"teléfono"      
"primera"       
"vez"
"cocinaron"     
"pregunta"      
"cantante"      
"Descubren"     
"ladrón"        
"cambia"        
"barren"        
"escoba"        
"recogió"       
"esposa"        
"caros"         
"padres"        
"limpia"        
"vestido"       
"baño"          
"firmó"         
"escaló"        
"adivina"       
"amigos"        
"investigan"    
"apagaron"      
"corta"         
"cepilló"       
"necesitó"      
"aprendió"      
"manda"         
"apaga"         
"salta"         
"cuerda"        
"gimnasio"      
"dibujaron"     
"corazón"       
"esposo"        
"hora"          
"clase"         
"grabó"         
"canceló"       
"facturaron"    
"recibe"        
"dibuja"        
"pajaro"        
"llevó"         
"parag¸as"      
"cancelan"      
"adivinaron"    
"gritó"         
"barre"         
"limpian"       
"mejora"        
"mujeres"       
"recogen"       
"comparte"      
"llenaron"      
"escalaron"     
"colina"        
"gastan"        
"imprimen"      
"levanta"       
"cortan"        
"del"
"cambió"        
"tía"
"escucha"       
"conversaciones"
"jefa"          
"ancianos"      
"aprenden"      
"imprimió"      
"abogado"       
"lleva"         
"llevaron"      
"despierta"     
"sala"          
"escuchó"       
"preguntó"      
"peluqueras"    
"cepillan"      
"jefes"         
"dibujó"        
"guardó"        
"esperó"        
"recibieron"    
"muchachos"     
"preguntan"     
"gastó"         
"saltó"         
"cárcel"        
"pájaro"        
"alemán"        
"paragüas"      
"novios"        
"nina"          
"ninas"         
"nino"          
"manana"        
"bano"          
"ano"
"companía"      
"móvil"         
"camareros"     
"montana"       
"ninos"         
"senora"        
"condicioens" 


```{r}

nsg <- c(
  "muchacho", "oficina"
)


npl <- c(
  "puros"
)


vpr <- c(
  "Cuentan", "fuma"
)

vpast <- c(
  "cortó"
)

det <- c(
  "el", "la"
)


adv <- c(
  "ahora"
)

adj <- c(
  "sus"
)

prep <- c(
  "que", "en"
)

clitics <- c(
  "lo"
)






data <- read_csv("./data/proficiency.csv") %>% 
  mutate(
    IA_LABEL = str_remove_all(IA_LABEL, " "), 
    IA_LABEL = str_remove_all(IA_LABEL, "\\."), 
    IA_LABEL = str_replace_all(IA_LABEL, "·", "a"), 
    IA_LABEL = str_replace_all(IA_LABEL, "Ì", "í"), 
    IA_LABEL = str_replace_all(IA_LABEL, "È", "é"), 
    IA_LABEL = str_replace_all(IA_LABEL, "Ò", "n"), 
    IA_LABEL = str_replace_all(IA_LABEL, "Û", "ó"), 
  ) %>% 
  mutate(
    word_type = case_when(
      IA_LABEL %in% nsg ~ "nsg", 
      IA_LABEL %in% npl ~ "npl", 
      IA_LABEL %in% vpr ~ "vpr", 
      IA_LABEL %in% vpast ~ "vpast", 
      IA_LABEL %in% det ~ "det", 
      IA_LABEL %in% adv ~ "adv", 
      IA_LABEL %in% adj ~ "adj", 
      IA_LABEL %in% prep ~ "prep", 
      IA_LABEL %in% clitics ~ "clitics", 
      TRUE ~ "NA"
    )
  )



glimpse(data)
```

```{r}

data$IA_LABEL %>% unique()
data$word_type



data %>%
  mutate(Word_Type = if_else(IA_LABEL == "Cuentan", "VPR", ""))

```